---
title: "NFL Combine Project"
author: "David Deniziuk"
date: "11/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Imports
```{r, warning=FALSE, message=FALSE}
#preemptively loading various libraries
library(readr)      
library(tidyverse)
library(dplyr)
library(tidyr)
library(leaps)
library(devtools)
library(data.table) 
```

```{r, warning=FALSE, message=FALSE}
data <- read_csv("combine_data.csv")  #reading in csv
```


# Quarterback Section
### Creation of Dataframe


predictor variables
```{r, warning=FALSE, message=FALSE}
#modification of dataset of predictor variables
data <- data %>% filter(Pos == "QB", Forty != "NA", Shuttle != "NA", BroadJump != "NA", Cone != "NA", Vertical != "NA", Team != "NA") #filtering for QB, as well as removing observations missing necessary predictor variables
data <- data %>%
  select(Player = Player, CombineYear = Year, Forty = Forty, Vertical = Vertical, BroadJump = BroadJump, Cone = Cone, Shuttle = Shuttle, Pick = Pick) # selected only the necessary predictor variables
head(data, 10)
```

response variables
```{r, warning=FALSE, message=FALSE}
#creation and modification of dataset of response variables
data_ <- read_csv("quarterbacks.csv")  #reading in csv
data_ <- data_ %>% filter(year >= 2000) %>% group_by(qb, year) %>% summarise(rate = mean(rate), games = length(qb)) %>% #filtered data for observations after year 2000 to align with the dataset of our predictor variables.  Creating response variable "rate" to be an average of the rate from each individual game in that season. 
  filter(games >= 10)   # filtered for seasons where QB played at least 10 games in order to select for starters

data_ %>% arrange(desc(year))
```

merging both dataframes
```{r}
#this code is designed to merge dataframes together. It was tricky due to the player names being listed differently in each dataset.  
data <- data.table(data, key = "Player")  #set up key to join 
data_ <- data.table(data_, key = "qb")  #set up key to join

temp <- data[,data_[agrep(Player, qb)], by = .(Player, CombineYear, Forty, Vertical, BroadJump, Cone, Shuttle, Pick)] #agrep function helps merge when keys (QB names) do not align perfectly
temp$qb <- NULL #removed duplicate column
```

final dataframe
```{r, warning=FALSE, message=FALSE}
#this code creates a temporary dataframe in order to get QBRs from a given QB's first year. 
firstyear_rate <- temp %>% 
  group_by(Player) %>% filter(year == min(year)) %>% #selects for minimum year
  arrange(Player)     #orders alphabetically
colnames(firstyear_rate)[10] <- "firstyearrate"  #rename variable
colnames(firstyear_rate)[9] <- "firstyear"  #rename variable


final <- temp %>% group_by(Player, CombineYear, Forty, Vertical, BroadJump, Cone, Shuttle) %>%  summarize(rate = mean(rate)) %>% #calculate mean of season QBRs for a career QBR.
  arrange(Player)
final$firstyearrate <- firstyear_rate$firstyearrate #add first year variable to dataframe
final$CombineYear <- NULL #remove unnecessary variable
```

```{r}
#final dataframe for modeling
head(final, 10)
```


### Quarterback Descriptive Statistics


descriptive statistics: predictor variables
```{r, warning=FALSE, message=FALSE}
attach(final)  #attach dataset for ease of use
mean(Forty)    #various means
mean(Vertical)
mean(BroadJump)
mean(Cone)
mean(Shuttle)

sd(Forty)    #various standard deviations
sd(Vertical)
sd(BroadJump)
sd(Cone)
sd(Shuttle)
```

```{r}
par(mfrow = c(2, 3))
hist(Forty)    #various histograms
hist(Vertical)
hist(BroadJump)
hist(Cone)
hist(Shuttle)
```

descriptive statistics: response variables
```{r, warning=FALSE, message=FALSE}
mean(rate)    #means
mean(firstyearrate)

sd(rate)    #standard deviations
sd(firstyearrate)

par(mfrow = c(2, 1))
hist(rate)    # histograms
hist(firstyearrate)
```

boxplot for career average QBR
```{r}
ggplot(final, aes(y=rate)) +   #creating boxplot
  geom_boxplot() +   
  coord_flip()+       #flipping coordinates
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())  #removing axis label
```

boxplot for first year only QBR
```{r}
ggplot(final, aes(y=firstyearrate)) +    #creating boxplot
  geom_boxplot() + 
  coord_flip()+     #flipping coordinates
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())  #removing axis label
#detach(final)
```


### Quarterback Modeling


fitting model 1: career average
```{r}
#fitting model on average QBR across years. "rate"
fit.full <- lm(rate ~ Forty+Vertical+BroadJump+Cone+Shuttle, data=final)  #fitting full model
fit.empty <- lm(rate~1, data=final)  #fitting empty model
fit.final <- step(fit.empty, direction = 'both', scope = formula(fit.full), trace = 0)  #iterating through stepwise selection
summary(fit.final) #printing summary of stepwise model
```

fitting model 2: first year only
```{r}
#fitting model on first year QBR. "firstyearrate"
fit.full_ <- lm(firstyearrate ~ Forty+Vertical+BroadJump+Cone+Shuttle, data=final)  #fitting full model
fit.empty_ <- lm(firstyearrate~1, data=final)  #fitting empty model
fit.final_ <- step(fit.empty_, direction = 'both', scope = formula(fit.full_), trace = 0)  #iterating through stepwise selection on previous model
summary(fit.final_) #printing summary of stepwise model
```


# Runningback Section
### Creation of Dataframe


predictor variables
```{r, warning=FALSE, message=FALSE}
#creation and modification of dataset of predictor variables
dataRB <- read_csv("combine_data.csv")  #reading in csv
dataRB <- dataRB %>% filter(Pos == "RB", Forty != "NA", Shuttle != "NA", BroadJump != "NA", Cone != "NA", Vertical != "NA", Team != "NA") #filtering for RB, as well as removing observations missing necessary predictor variables
dataRB <- dataRB %>%
  select(Player = Player, CombineYear = Year, Forty = Forty, Vertical = Vertical, BroadJump = BroadJump, Cone = Cone, Shuttle = Shuttle, Pick = Pick) # selected only the necessary predictor variables
dataRB %>% arrange(desc(CombineYear)) #displaying dataframe ordered by combine year
```

response variables
```{r, warning=FALSE, message=FALSE}
#creation and modification of dataset of response variables
dataRB_ <- read_csv("runningbacks.csv")  #reading in csv
dataRB_
colnames(dataRB_)[16] <- "YardsPerCarry"
dataRB_$YardsPerCarry <- as.numeric(dataRB_$YardsPerCarry) 

dataRB_ <- dataRB_ %>% filter(Year >= 2000) %>% filter(Season == "Regular Season") %>% drop_na(YardsPerCarry) %>% group_by(`Player Id`, Year) %>%   summarise(YardsPerCarry = mean(YardsPerCarry), games = length(`Player Id`)) %>% #filtered data for observations after year 2000 to align with the dataset of our predictor variables.  Selected for regular season games.  Removed NA values of our response variable, Yards Per Carry.  Created new, meaningful variables 
     filter(games >= 10)   # filtered for seasons where RB played at least 10 games in order to select for starters
   
dataRB_ %>% arrange(desc(Year))  #displaying dataframe ordered by year
```
 
merging both data frames
```{r}
#this code is designed to merge dataframes together. It was tricky due to the player names being listed differently in each dataset.  
dataRB <- data.table(dataRB, key = "Player")  #set up key so join works
dataRB_ <- data.table(dataRB_, key = "Player Id")  #set up key so join works

tempRB <- dataRB[,dataRB_[agrep(Player, `Player Id`, max.distance = 0.35)], by = .(Player, CombineYear, Forty, Vertical, BroadJump, Cone, Shuttle)] #agrep function helps merge when names do not align perfectly
tempRB <- tempRB %>% filter(Player != "Adrian Peterson-02") #removed duplicate Adrian Peterson lol
tempRB
``` 
 
final dataframe
```{r, warning=FALSE, message=FALSE}
#this code creates a temporary dataframe in order to get yards per carry from a given RB's first year. 
firstyear_YPC <- tempRB %>% 
  group_by(Player) %>% filter(Year == min(Year)) %>% #selects for minimum year
  arrange(Player)     #orders alphabetically
colnames(firstyear_YPC)[10] <- "firstyearYardsPerCarry"  #rename variable
colnames(firstyear_YPC)[9] <- "firstyear"  #rename variable
firstyear_YPC <- firstyear_YPC %>% filter(Player != "Dan Herron") #remove mismatched observation 


finalRB <- tempRB %>% group_by(Player, CombineYear, Forty, Vertical, BroadJump, Cone, Shuttle) %>%  summarize(YardsPerCarry = mean(YardsPerCarry)) %>% filter(Player != "Dan Herron") %>% arrange(Player) #calculate mean of season YPCs, remove mismatched observation  
finalRB$firstyearYardsPerCarry <- firstyear_YPC$firstyearYardsPerCarry #add first year variable to dataframe
finalRB$CombineYear <- NULL #remove unnecessary variable

```

```{r}
#final dataframe for modeling
#finalRB
```


### Runningback Descriptive Statistics


descriptive statistics: predictor variables
```{r, warning=FALSE, message=FALSE}
attach(finalRB)  #attach dataset for ease of use
mean(Forty)    #various means
mean(Vertical)
mean(BroadJump)
mean(Cone)
mean(Shuttle)

sd(Forty)    #various standard deviations
sd(Vertical)
sd(BroadJump)
sd(Cone)
sd(Shuttle)
```

```{r}
par(mfrow = c(2, 3))
hist(Forty)    #various histograms
hist(Vertical)
hist(BroadJump)
hist(Cone)
hist(Shuttle)
```

descriptive statistics: response variables
```{r, warning=FALSE, message=FALSE}
mean(YardsPerCarry)    #means
mean(firstyearYardsPerCarry)

sd(YardsPerCarry)    #standard deviations
sd(firstyearYardsPerCarry)

par(mfrow = c(2, 1))
hist(YardsPerCarry)    # histograms
hist(firstyearYardsPerCarry)
```

boxplot for career average YPC
```{r}
ggplot(finalRB, aes(y=YardsPerCarry)) +   #creating boxplot
  geom_boxplot() +   
  coord_flip()+       #flipping coordinates
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())  #removing axis label
```

boxplot for first year only YPC
```{r}
ggplot(finalRB, aes(y=firstyearYardsPerCarry)) +    #creating boxplot
  geom_boxplot() + 
  coord_flip()+     #flipping coordinates
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())  #removing axis label
detach(finalRB)
```


### Runningback Modeling


fitting model 1: career average
```{r}
#fitting model on average Yards Per Carry across years. "YardsPerCarry"
fit.fullRB <- lm(YardsPerCarry ~ Forty+Vertical+BroadJump+Cone+Shuttle, data=finalRB)  #fitting full model
fit.emptyRB <- lm(YardsPerCarry~1, data=finalRB)  #fitting empty model
fit.finalRB <- step(fit.emptyRB, direction = 'both', scope = formula(fit.fullRB), trace = 0)  #iterating through stepwise selection on previous model
summary(fit.finalRB)
```

fitting model 2: first year only
```{r}
#fitting model on first year YPC. "firstyearYardsPerCarry"
fit.fullRB_ <- lm(firstyearYardsPerCarry ~ Forty+Vertical+BroadJump+Cone+Shuttle, data=finalRB)  #fitting full model
fit.emptyRB_ <- lm(firstyearYardsPerCarry~1, data=finalRB)  #fitting empty model
fit.finalRB_ <- step(fit.emptyRB_, direction = 'both', scope = formula(fit.fullRB_), trace = 0)  #iterating through stepwise selection on previous model
summary(fit.finalRB_)
```

